---
title: "project"
author: ""
date: "2025-11-25"
output: html_document
---

The dataset used in this project corresponds to the supplementary file **MOESM2 (ESM)** from the publication by Fan et al. (2018). It contains the raw GC-TOF-MS metabolomics data for 120 urine samples\
from healthy adults (60 males and 60 females), enabling independent preprocessing, exploratory analysis, feature selection, and classification.

# Import Library

```{r}
library(readxl)
library(tidyverse)
library(ggplot2)
library(ellipse)
library(caret)
library(pROC)
library(pls)
library(DescTools)
library(dplyr)
library(mixOmics)
```

# Import dataset

```{r}
raw_df <- read_excel("41598_2018_29592_MOESM2_ESM.xlsx")
```

```{r}
sex <- unlist(raw_df[3, -1])
sex <- as.factor(sex)

df <- read_excel("41598_2018_29592_MOESM2_ESM.xlsx", skip = 6)

glimpse(df)
```

# Cleaning and Data Transformation

```{r}
# organize names
# first df line has de colnames
colnames(df) <- df[1, ]

df <- df[-1, ]

# convert to numeric
for (i in 17:ncol(df)) {  # first 16 columns = metadata
  df[[i]] <- as.numeric(df[[i]])
}

X <- df[, 17:ncol(df)] 

Y <- sex

colnames(X) <- paste0("S", seq_len(ncol(X)))

# number of metabolites and samples
dim(X)

```

It matches with the article 120 samples (60 males + 60 females)

# Summary statistics

```{r}
summary(X)
```

```{r}
# Count total zero values
sum(X == 0, na.rm = TRUE)
```

```{r}
# Missing values check
sum(is.na(X))
```

# Is it the data Gaussian Distributed?

```{r}
par(mfrow=c(1,2))
hist(as.numeric(X[1,]), main="Raw Distribution", xlab="Intensity")
qqnorm(as.numeric(X[1,]), main="QQ Plot Raw")
qqline(as.numeric(X[1,]))
par(mfrow=c(1,1))
```

# Non linear

```{r}
# Convert to standard data frame to avoid tibble recycling issues
X_nozero <- as.data.frame(X)

for(i in 1:nrow(X_nozero)){
  row_vals <- as.numeric(X_nozero[i, ])  # convert row to numeric vector

  # Safeguard: if the row is all zeros (rare but possible)
  if(all(row_vals == 0)){
    row_vals[row_vals == 0] <- 1  # placeholder before log10
  } else {
    min_val <- min(row_vals[row_vals > 0], na.rm = TRUE)
    row_vals[row_vals == 0] <- min_val / 2
  }

  X_nozero[i, ] <- row_vals  # assign back properly
}
```

The matrix was transposed so that samples correspond to rows and metabolites to columns, which is the correct structure required for multivariate analyses such as PCA and PLS-DA, where each row must represent an independent observation:

-   With X_scaled \<- scale(t(X_log)).

```{r}
# Log10 transformation
X_log <- log10(X_nozero)

# Scaling for PCA (samples as rows)
X_scaled <- scale(t(X_log))

# PCA
pca <- prcomp(X_scaled, center = TRUE, scale. = FALSE)
summary(pca)

```

-   Summary statistics show very different scales and non-normal distributions

-   A few zero values detected → replaced with half of minimum value\
    → required for log10 transformation

-   Applied log10(x) to reduce skewness and stabilize variance

-   Autoscaling (mean-center + unit variance) to make metabolites comparable

```{r}
sex_row <- as.character(raw_df[3, ])

sample_names_raw <- sex_row[sex_row %in% c("Male", "Female")]

sample_cols <- colnames(X)

sex <- factor(sample_names_raw[1:length(sample_cols)])

length(sex)
table(sex)

all(!is.na(sex))
identical(length(sex), nrow(pca$x))

```

```{r}
pca_scores <- as.data.frame(pca$x[, 1:2])


center <- colMeans(pca_scores)
cov_mat <- cov(pca_scores)

distances <- mahalanobis(pca_scores, center, cov_mat)

# Cutoff (chi with p=0.95 and 2 df)
cutoff <- qchisq(0.95, df = 2)

# plot with condidese matrix
plot(pca$x[,1], pca$x[,2], 
     pch = 19, 
     col = ifelse(distances > cutoff, "red", "black"), # Outliers a vermelho
     xlab = "PC1", ylab = "PC2", 
     main = "PCA Outlier Detection (95% CI)")

lines(ellipse(cov_mat, centre = center, level = 0.95), col="blue", lty=2)

text(pca$x[,1], pca$x[,2], 
     labels = ifelse(distances > cutoff, rownames(pca$x), ""), 
     pos = 3, cex = 0.7)

# Identify outliers
outliers <- which(distances > cutoff)
print(paste("Outliers detected:", length(outliers)))
```

We detected 9 outliers, but they weren't removed because:

Outliers reflect **real biological variability** within the healthy population.\
Removing them could introduce **bias**, especially if outliers are unevenly distributed between sexes.\
Furthermore, the original study **did not exclude any samples** all **120 subjects** were analyzed in full.

```{r}
score_df <- data.frame(
  PC1 = pca$x[,1],
  PC2 = pca$x[,2],
  Sex = sex
)

ggplot(score_df, aes(PC1, PC2, color = Sex)) +
  geom_point(size = 3) +
  theme_classic() +
  ggtitle("PCA After Log10 Transform + Scaling")


```

```{r}
# Scree plot (PCA Variance Explained)
variance <- pca$sdev^2 / sum(pca$sdev^2)

plot(variance[1:10], type="b", pch=19,
     xlab="PC", ylab="Proportion of Variance",
     main="Scree Plot - PCA")

```

-   Score plot shows a trend of separation between Male and Female

-   Some possible outliers appear (next step: detection)

-   Scree Plot indicates PC1+PC2 explain \~20–30% variance → normal for metabolomics

```{r}
set.seed(123)

train_index <- createDataPartition(sex, p = 0.7, list = FALSE)

X_train <- X_scaled[train_index, ]
X_test  <- X_scaled[-train_index, ]

X_train <- as.data.frame(X_train)
X_test  <- as.data.frame(X_test)

y_train <- sex[train_index]
y_test  <- sex[-train_index]

pvals <- apply(X_train, 2, function(x) t.test(x ~ y_train)$p.value)
pvals_fdr <- p.adjust(pvals, method = "fdr")

head(sort(pvals_fdr))


```

```{r}
#Initial PLS-DA model
pls_initial <- plsda(X_train, y_train, ncomp = 5)

plotIndiv(pls_initial, comp = c(1,2),
          group = y_train, legend = TRUE,
          title = "Initial PLS-DA (Training Set)")
```

Internal Validation: 5-Fold CV + AUCROC

```{r}

set.seed(123)

folds <- createFolds(y_train, k = 5, returnTrain = TRUE)
auc_results <- data.frame(LV = integer(), AUC = numeric())

for (lv in 1:5) {
  auc_fold <- c()
  
  for (f in folds) {
    model <- plsda(X_train[f,], y_train[f], ncomp = lv)
    pred <- predict(model, X_train[-f,])$predict[,1,lv]
    roc_obj <- roc(y_train[-f], pred)
    auc_fold <- c(auc_fold, auc(roc_obj))
  }
  
  auc_results <- rbind(auc_results,
                       data.frame(LV = lv,
                                  AUC = mean(auc_fold)))
}

auc_results

```

AUC improves until LV = 3, then decreases slightly. Therefore, LV=3 is the optimal number of components.

Univariate Analysis on Training set (Male vs Female)

```{r}
# Compute raw p-values using Welch t-test for each metabolite
pvals <- apply(X_train, 2, function(x) t.test(x ~ y_train)$p.value)

# FDR correction (Benjamini-Hochberg)
pvals_fdr <- p.adjust(pvals, method = "fdr")


fc <- apply(X_train, 2, function(x) mean(x[y_train == "Male"]) -
                               mean(x[y_train == "Female"]))


uni_results <- data.frame(
  Metabolite = colnames(X_train),
  p_value = pvals,
  p_FDR = pvals_fdr,
  mean_diff = fc
)


uni_results <- uni_results[order(uni_results$p_FDR), ]
head(uni_results)

```

```{r}
significant <- uni_results[uni_results$p_FDR < 0.05, ]
significant

```

```{r}
library(ggplot2)

volcano <- uni_results
volcano$logP <- -log10(volcano$p_FDR)

ggplot(volcano, aes(x = mean_diff, y = logP)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = -log10(0.05), col = "red", lty = 2) +
  theme_classic() +
  xlab("Mean Difference (Male – Female)") +
  ylab("-log10(FDR)") +
  ggtitle("Volcano Plot — Univariate Analysis (Training Set)")

```

Univariate analysis performed on the training set identified 9 metabolites with significant sex-related differences after FDR correction (p_FDR \< 0.05). Six metabolites (V22, V399, V84, V51, V291, V57) showed higher concentrations in females, while three metabolites (V10, V113, V250) were higher in males. The strongest discriminators were V22 and V399 (p_FDR \< 0.005).

